---
title: "`bayesmsm` for longitudinal data without right-censoring"
author: "Xiao Yan, Kuan Liu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`bayesmsm` for longitudinal data without right-censoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = 'center',
  fig.width = 9,
  fig.height = 6,
  warning = F,
  message = F
)
```

# Introduction

- The `bayesmsm` package enables easy implementation of the Bayesian marginal structural models (BMSMs) for longitudinal data. The methodology of BMSMs can be divided into 2 estimation steps:

  - Step 1. Bayesian treatment effect weight estimation
  - Step 2. Bayesian non-parametric bootstrap to maximize the utility function with respect to the causal effect
  
- For Step 1, we estimate treatment weights $w_{ij}$ using posterior samples of the $\alpha$ and $\beta$ via fitting a series of logistic regressions in a Bayesian framework. The package incorporates both Inverse Probability of Treatment Weighting (IPTW) and Inverse Probability of Censoring Weighting (IPCW) to handle longitudinal data without and with right-censoring. For Step 2, $P_n(v_{ij})$ is estimated via non-parametric Bayesian bootstrap with $Dir(1,...,1)$ sampling weights.

- The main functions in this package include:

  - `bayesweight`: Calculates Bayesian weights for subject-specific treatment effects.
  - `bayesweight_cen`: Calculates Bayesian weights for subject-specific treatment effects with right-censored data.
  - `bayesmsm`: Estimates marginal structural models using the calculated Bayesian weights.
  - `plot_ATE`: Plots the estimated Average Treatment Effect (ATE).
  - `plot_APO`: Plots the estimated Average Potential Outcome (APO).
  - `plot_est_box`: Plots the distribution of estimated treatment effects.
  - `summary_bayesmsm`: Summarizes the model results from `bayesmsm`.

- Installation

  - To install the bayesmsm package, you can use the `devtools` package to install it directly from GitHub:

```r
devtools::install_github("Kuan-Liu-Lab/bayesmsm")
library(bayesmsm)
```

```{r, warning = FALSE, results=FALSE, force=TRUE, include=FALSE}
if (!requireNamespace("bayesmsm", quietly = TRUE)) {
  stop("The package 'bayesmsm' is required to run this vignette. Please install it manually using:
       devtools::install_github('Kuan-Liu-Lab/bayesmsm')")
} else {
  library(bayesmsm)
}
```

# Simulated observational data with a time-dependent treatment

- The simulated dataset (continuous outcome)
  - 1000 patients and 3 visits (2 of which patients were assigned a treatment)
  - y, an end-of-study continuous outcome
  - z, a binary treatment
  - w1 and w2 are two baseline covariates (one continuous and one binary, mimicking age and sex)
  - L1 and L2 are two time-dependent covariates (also one continuous and one binary)
  - no missing data


```{r}
library(DT)

testdata <- read.csv(system.file("extdata", "continuous_outcome_data.csv", package = "bayesmsm"))

# look at the data;
datatable(testdata,
          rownames = FALSE,
          options = list(dom = 't')) %>%
  formatRound(columns=c('w2', 'L2_1', 'L2_2', 'y'), digits=2)
```

- Frequency Counts by Treatment Combinations

```{r}
# frequency counts by treatment combinations;
table(testdata$a_1, testdata$a_2)
```

- Suppose the causal parameter of interest is the average treatment effect between always treated and never treated,

$$
ATE = E(Y \mid Z_1 = 1, Z_2 = 1) - E(Y \mid Z_1 = 0, Z_2 = 0)
$$

# Bayesian treatment effect weight estimation using `bayesweight`

- The following code calls the function `bayesweight` to run JAGS and calculate the weights.
  - Non-parallel computing requires that `n.chains = 1`. Parallel MCMC requires at least 2 chains because computing is running on 1 core per chain, and we recommend using at most 2 chains less than the number of available cores on your computer.
  - Running this function automatically saves a JAGS model file in the working directory, which the user can check to review the model specifications.

- Parameters Description:
  - `trtmodel.list`: A list of formulas corresponding to each time point with the time-specific treatment variable on the left hand side and pre-treatment covariates to be balanced on the right hand side. Interactions and functions of covariates are allowed.
  - `data`: The dataset containing all the variables specified in trtmodel.list.
  - `n.chains`: Number of MCMC chains to run. For non-parallel execution, this should be set to 1. For parallel execution, it requires at least 2 chains.
  - `n.iter`: Total number of iterations for each chain (including burn-in).
  - `n.burnin`: Number of iterations to discard at the beginning of the simulation (burn-in).
  - `n.thin`: Thinning rate for the MCMC sampler.
  - `seed`: Seed to ensure reproducibility.
  - `parallel`: Logical flag indicating whether to run the MCMC chains in parallel. Default is TRUE.

```{r}
weights <- bayesweight(trtmodel.list = list(a_1 ~ w1 + w2 + L1_1 + L2_1,
                                             a_2 ~ w1 + w2 + L1_1 + L2_1 + L1_2 + L2_2 + a_1),
                        data = testdata,
                        n.chains = 1,
                        n.iter = 200,
                        n.burnin = 100,
                        n.thin = 1,
                        seed = 890123,
                        parallel = FALSE)
str(weights)
```

- It returns a list containing:
  - `weights`: The calculated weights for subject-specific treatment effects, computed by taking the average of the weights across all MCMC iterations.
  - `model_string`: A character of the JAGS model based on input of the argument trtmodel.list.

- We can print and view the JAGS model stored in `model_string` from the above function output:

```{r}
cat(weights$model_string)
```


# Bayesian non-parametric bootstrap to maximize the utility function with respect to the causal effect using `bayesmsm`

The function `bayesmsm` estimates causal effect of time-varying treatments. It uses subject-specific treatment assignment weights *weights* calculated using `bayesweight`, and performs Bayesian non-parametric bootstrap to estimate the causal parameters.

- Parameters Description:
  - `ymodel`: A formula representing the outcome model, which can include interactions and functions of covariates.
  - `nvisit`: Specifies the number of visits or time points considered in the model.
  - `reference`: The baseline or reference intervention across all visits, typically represented by a vector of zeros indicating no treatment (default is a vector of all zeros).
  - `comparator`: The comparison intervention across all visits, typically represented by a vector of ones indicating full treatment (default is a vector of all ones).
  - `family`: Specifies the outcome distribution family; use "gaussian" for continuous outcomes or "binomial" for binary outcomes (default is "gaussian").
  - `data`: The dataset containing all variables required for the model.
  - `wmean`: A vector of treatment assignment weights. Default is a vector of ones, implying equal weighting.
  - `nboot`: The number of bootstrap iterations to perform for estimating the uncertainty around the causal estimates.
  - `optim_method`: The optimization method used to find the best parameters in the model (default is 'BFGS').
  - `seed`: A seed value to ensure reproducibility of results.
  - `parallel`: A logical flag indicating whether to perform computations in parallel (default is TRUE).
  - `ncore`: The number of cores to use for parallel computation (default is 4).

```{r}
model <- bayesmsm(ymodel = y ~ a_1+a_2,
                           nvisit = 2,
                           reference = c(rep(0,2)),
                           comparator = c(rep(1,2)),
                           family = "gaussian",
                           data = testdata,
                           wmean = weights$weights,
                           nboot = 50,
                           optim_method = "BFGS",
                           parallel = FALSE,
                           seed = 890123,
                           ncore = 2)
str(model)
```

- It returns a model object which contains:
  - `mean`, `sd`, `quantile`: the mean, standard deviation and 95% credible interval of the estimated causal effect (ATE). From the above results, the mean of ATE is approximately -3.161, which indicates that the expected outcome for always treated patients is, on average, 3.161 units less than that for never treated patients. 
  - `bootdata`: a data frame containing the bootstrap samples for the reference effect, comparator effect, and average treatment effect (ATE).
  - `reference`, `comparator`: the reference level and comparator level the user chooses to compare. Here the reference level is never treated (0,0), and the comparator level is always treated (1,1).

# Visualization functions: `plot_ATE`, `plot_APO`, `plot_est_box`

The `bayesmsm` package also provides several other functions for visualizing the above results: `plot_ATE`, `plot_APO`, and `plot_est_box`. These functions help the user better interpret the estimated causal effects.

- Plotting the Average Treatment Effect (ATE)
  - The `plot_ATE` function generates a plot of the estimated ATE with its 95% credible interval.
  
```{r}
plot_ATE(model)
```

- Plotting the Average Potential Outcome (APO)
  - Similarly, the `plot_APO` function plots the estimated APO for both the reference and comparator level effects.

```{r}
plot_APO(model, "effect_reference")
plot_APO(model, "effect_comparator")
```

- Plotting the Distribution of Estimated Treatment Effects
  - The `plot_est_box` function generates an error bar plot of the estimated treatment effects (APO and ATE) from the bootstrap samples.

```{r}
plot_est_box(model)
```

# Reference

- Liu, K. (2021). Bayesian causal inference with longitudinal data. Tspace.library.utoronto.ca. https://tspace.library.utoronto.ca/handle/1807/109330
- Saarela, O., Stephens, D. A., Moodie, E. E. M., & Klein, M. B. (2015). On Bayesian estimation of marginal structural models. Biometrics, 71(2), 279–288. https://doi.org/10.1111/biom.12269
- Robins, J. M., Hernán, M. A., & Brumback, B. (2000). Marginal structural models and causal inference in epidemiology. Epidemiology, 11(5), 550–560. https://doi.org/10.1097/00001648-200009000-00011
- Liu, K., Saarela, O., Feldman, B. M., & Pullenayegum, E. (2020). Estimation of causal effects with repeatedly measured outcomes in a Bayesian framework. Statistical Methods in Medical Research, 29(9), 2507–2519. https://doi.org/10.1177/0962280219900362
